Concretizing the Process Hitting into Biological Regulatory Networks

### 2013-04-17 parts/infer-param.tex
Add TODOs and REWRITEs
add unsigned edges to parameters inference

This subsection presents some results related to the inference of independent discrete parameters from a given PH. These results are equivalent to those presented in \cite{PMR10-TCSB}, with notation adapted to be shared with the previous section.
In addition, we introduce the well-formed PH for parameter inference property (\pref{pro:wf-ph-K}),
which implies that the inferred IG does not contain any unsigned interactions, and thus can be seen as the
regular IG $(\Gamma, E_+, E_-)$,
and that any processes in $\levelsA{b}{a}$ (resp. $\levelsI{b}{a}$) share the same behavior
regarding $a$.
\begin{property}[Well-formed PH for parameter inference]\label{pro:wf-ph-K}
A PH is well-formed for parameter inference if and only if
it is well-formed for IG inference, and
the IG $(\Gamma, E_+, E_-, E_\pm)$ inferred by \pref{pps:inference-IG}
verifies $E_\pm=\emptyset$ and if the following property holds:
\begin{align}
\begin{split}
\forall b\in \GRNreg{a} &,
	\forall (i,j\in\levelsA{b}{a} \vee i,j\in\levelsI{b}{a}), \\
& \qquad	\forall c, ( (b\neq a\wedge c=a) \vee (c\in\PHpredec{a} \wedge b\in\PHdirectpredec{c})), \\
& \qquad \qquad
			\PHfrappe{b_i}{c_k}{c_l}\in\PHa \Leftrightarrow
				\PHfrappe{b_j}{c_k}{c_l}\in\PHa
\end{split}
\label{eq:wf-levels}
\end{align}
\end{property}

Let $K_{a,A,B}$ be the parameter we want to infer for a given component $a \in \Gamma$
%and $A,B \in \GRNallres{a}$ a configuration of resources of $a$ (activators and inhibitors).
and $A \subset \GRNreg{a}$ (resp. $B \subset \GRNreg{a}$) a set of its activators (resp. inhibitors).
This inference, as for the IG inference, relies on the search of focal processes of the component for the given configuration of its regulators.

For each sort $b \in \GRNreg{a}$, we define a context $C^b_{a,A,B}$ in \pref{eq:param_context} that contains all processes representing the influence of the regulators in the configuration $A,B$.
The context of a cooperative sort $\upsilon$ that regulates $a$ is given in
\pref{eq:param_context_coop} as the set of focal processes matching the current configuration.
$C_{a,A,B}$ refers to the union of all these contexts (\pref{eq:K-ctx}).
\begin{align}
\label{eq:param_context}
\forall b\in\Gamma,~
C_{a,A,B}^b & \DEF \begin{cases}
\levelsA{b}{a} & \text{if $b \in A$,}\\
\levelsI{b}{a} & \text{if $b \in B$,}\\
L_b		& \text{otherwise;}\\
\end{cases}
\\
\label{eq:param_context_coop}
\forall \upsilon \in \PHpredec{a}\setminus\Gamma,~
C_{a,A,B}^\upsilon & \DEF \{
\upsilon(\sigma) \mid \sigma \in \textstyle\prod_{c\in\PHdirectpredec{\upsilon}}C_{a,A,B}^c \}
\\
C_{a,A,B} & \DEF \textstyle\bigcup_{b\in\PHpredec{a}} C^b_{a,A,B}
\label{eq:K-ctx}
\end{align}

The parameter $K_{a,A,B}$ specifies to which values $a$ eventually evolves as long as the context
$C_{a,A,B}$ holds, which is precisely the definition of the $\focals$ function
(\pref{def:focals} in \pref{ssec:focal}),
where the focals reachability property can be derived from \pref{pro:wf-ph-K} and
\pref{eq:param_context_coop}.
Hence $K_{a,A,B} = \focals(a,C^a_{a,A,B},C_{a,A,B})$ if this latter is a non-empty interval
(\pref{pps:param_K}).

\begin{proposition}[Parameter inference]
\label{pps:param_K}
Let $(\PHs, \PHl, \PHh)$ be a Process Hitting well-formed for parameter inference, and $\IG = (\Gamma,
E_+, E_-)$ the inferred IG.
Let $A$ (resp. $B$) $\subseteq \Gamma$ be the set of regulators that activate (resp. inhibit) a sort
$a$.
%If $\focals(a,C_{a,A,B})$ is a non-empty interval, then $K_{a,A,B} = \focals(a, C_{a,A,B})$.
If $\focals(a,C^a_{a,A,B},C_{a,A,B})=[a_i;a_j]$ is a non-empty interval, 
	then $K_{a,A,B} = [i;j]$.
\end{proposition}

@@

\begin{property}[Parameter validity]\label{pro:K-valid}
A parameter $K_{a,A,B}$ is valid w.r.t. the PH iff the following equation is verified:
\begin{align*}
\forall a_i\in C^a_{a,A,B},
		a_i \notin K_{a,A,B} \Longrightarrow (
  \exists \PHfrappe{c_k}{a_i}{a_j}\in\PHa, & c_k \in C^c_{a,A,B} \\
 \wedge a_i < K_{a,A,B} \Rightarrow j > i 
 & \wedge  a_i > K_{a,A,B} \Rightarrow j <i )
\end{align*}
\end{property}

@@

\begin{property}[Extreme values assumption]
Let $\IG = (\Gamma, E_+, E_-)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{extreme values assumption} iff:
\label{prop:param_enum_extreme}
\[
  \forall b \in \Gamma, \GRNreg{b} \neq \emptyset \Rightarrow 0 \in K_{b,\emptyset,\GRNreg{b}} \wedge l_b \in K_{b,\GRNreg{b},\emptyset}
\]
\end{property}

\begin{property}[Activity assumption]
\label{prop:param_enum_activity}
Let $\IG = (\Gamma, E_+, E_-)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{activity assumption} iff:
\begin{align*}
  \forall b \in \Gamma, \forall a \in \GRNreg{b}, \exists (A;B) \in \GRNallres{a}, K_{b,A,B} <_{[]} K_{b,A \cup \{b\},B \setminus \{b\}}
\\
  \forall b \in \Gamma, \forall a \in \GRNreg{b}, \exists (A;B) \in \GRNallres{a}, K_{b,A \setminus \{b\},B \cup \{b\}} <_{[]} K_{b,A,B}
\end{align*}
\end{property}

\begin{property}[Monotonicity assumption]
\label{prop:param_enum_monotonicity}
Let $\IG = (\Gamma, E_+, E_-)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{monotonicity assumption} iff:
\[
  \forall b \in \Gamma, \forall (A;B), (A';B') \in \GRNallres{b},
  A \subset A' \wedge B' \subset B \Rightarrow K_{b,A,B} \leq_{[]} K_{b,A',B'}
\]
\end{property}

@@

>>>>

\todo{Try to address the inference of parameters despite unsigned edges}

This subsection presents some results related to the inference of independent discrete parameters from a given PH. These results are equivalent to those presented in \cite{PMR10-TCSB}, with notation adapted to be shared with the previous section.
In addition, we introduce the well-formed PH for parameter inference property (\pref{pro:wf-ph-K}),
which implies that the inferred IG does not contain any unsigned interactions, and thus can be seen as the
regular IG $(\Gamma, E_+, E_-)$,
and that any processes in $\levelsA{b}{a}$ (resp. $\levelsI{b}{a}$) share the same behavior
regarding $a$.
\begin{property}[\new{Well-formed PH for parameter inference}]\label{pro:wf-ph-K}
A PH is well-formed for parameter inference if and only if
it is well-formed for IG inference, and
the IG $(\Gamma, E_+, E_-, E_\pm)$ inferred by \pref{pps:inference-IG}
verifies $E_\pm=\emptyset$ and if the following property holds:
\begin{align}
\begin{split}
\forall b\in \GRNreg{a} &,
	\forall (i,j\in\levels{b}{a} \vee i,j\in\ulevels{b}{a}), \\
& \qquad	\forall c, ( (b\neq a\wedge c=a) \vee (c\in\PHpredec{a} \wedge b\in\PHdirectpredec{c})), \\
& \qquad \qquad
			\PHfrappe{b_i}{c_k}{c_l}\in\PHa \Leftrightarrow
				\PHfrappe{b_j}{c_k}{c_l}\in\PHa
\end{split}
\label{eq:wf-levels}
\end{align}
\end{property}

Let $K_{a,\omega}$ be the parameter we want to infer for a given component $a \in \Gamma$
%and $A,B \in \GRNallres{a}$ a configuration of resources of $a$ (activators and inhibitors).
and $\omega \subset \GRNreg{a}$ a set of its regulators.
This inference, as for the IG inference, relies on the search of focal processes of the component for the given configuration of its regulators.

For each sort $b \in \GRNreg{a}$, we define a context $C^b_{a,\omega}$ in \pref{eq:param_context} that contains all processes representing the influence of the resources in the configuration modeled by $\omega$.
The context of a cooperative sort $\upsilon$ that regulates $a$ is given in
\pref{eq:param_context_coop} as the set of focal processes matching the current configuration.
$C_{a,\omega}$ refers to the union of all these contexts (\pref{eq:K-ctx}).
\begin{align}
\label{eq:param_context}
\forall b\in\Gamma,~
C_{a,\omega}^b & \DEF \begin{cases}
  \levels{b}{a} & \text{if $b \in \omega$,}\\
  \ulevels{b}{a} & \text{if $b \notin \omega$,}\\
  L_b            & \text{otherwise;}\\
\end{cases}
\\
\label{eq:param_context_coop}
\forall \upsilon \in \PHpredec{a}\setminus\Gamma,~
C_{a,\omega}^\upsilon & \DEF \{
\upsilon(\sigma) \mid \sigma \in \textstyle\prod_{c\in\PHdirectpredec{\upsilon}}C_{a,\omega}^c \}
\\
C_{a,\omega} & \DEF \textstyle\bigcup_{b\in\PHpredec{a}} C^b_{a,\omega}
\label{eq:K-ctx}
\end{align}

The parameter $K_{a,\omega}$ specifies to which values $a$ eventually evolves as long as the context
$C_{a,\omega}$ holds, which is precisely the definition of the $\focals$ function
(\pref{def:focals} in \pref{ssec:focal}),
where the focals reachability property can be derived from \pref{pro:wf-ph-K} and
\pref{eq:param_context_coop}.
Hence $K_{a,\omega} = \focals(a,C^a_{a,\omega},C_{a,\omega})$ if this latter is a non-empty interval
(\pref{pps:param_K}).

\begin{proposition}[Parameter inference]
\label{pps:param_K}
Let $(\PHs, \PHl, \PHh)$ be a Process Hitting well-formed for parameter inference, and $\IG = (\Gamma, E)$ the inferred IG.
Let $A$ (resp. $B$) $\subseteq \Gamma$ be the set of regulators that activate (resp. inhibit) a sort
$a$.
%If $\focals(a,C_{a,A,B})$ is a non-empty interval, then $K_{a,A,B} = \focals(a, C_{a,A,B})$.
If $\focals(a,C^a_{a,\omega},C_{a,\omega})=[a_i;a_j]$ is a non-empty interval, 
	then $K_{a,\omega} = [i;j]$.
\end{proposition}

@@

\begin{property}[Parameter validity]\label{pro:K-valid}
A parameter $K_{a,\omega}$ is valid w.r.t. the PH iff the following equation is verified:
\begin{align*}
\forall a_i\in C^a_{a,\omega},
		a_i \notin K_{a,\omega} \Longrightarrow (
  \exists \PHfrappe{c_k}{a_i}{a_j}\in\PHa, & c_k \in C^c_{a,\omega} \\
 \wedge a_i < K_{a,\omega} \Rightarrow j > i 
 & \wedge  a_i > K_{a,\omega} \Rightarrow j <i )
\end{align*}
\end{property}

@@

\begin{property}[Extreme values assumption]
Let $\IG = (\Gamma, E)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{extreme values assumption} iff:
\label{prop:param_enum_extreme}
\[
  \forall b \in \Gamma, \GRNreg{b} \neq \emptyset \Longrightarrow \exists \omega \subset \GRNreg{b}, 0 \in K_{b,\omega} \wedge \exists \omega' \subset \GRNreg{b}, l_b \in K_{b,\omega'}
\]
\end{property}

\begin{property}[Activity assumption]
\label{prop:param_enum_activity}
Let $\IG = (\Gamma, E)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{activity assumption} iff:
\begin{align*}
%  \forall b \in \Gamma, \forall a \in \GRNreg{b}, \exists (A;B) \in \GRNallres{a}, K_{b,A,B} <_{[]} K_{b,A \cup \{b\},B \setminus \{b\}}
%\\
%  \forall b \in \Gamma, \forall a \in \GRNreg{b}, \exists (A;B) \in \GRNallres{a}, K_{b,A \setminus \{b\},B \cup \{b\}} <_{[]} K_{b,A,B}
  \forall b \in \Gamma, \forall a \in \GRNreg{b}, \exists \omega \subset \GRNreg{b}, K_{b,\omega} \neq K_{b,\omega \cup \{ a \}}
\end{align*}
\end{property}

\begin{property}[Monotonicity assumption]
\label{prop:param_enum_monotonicity}
Let $\IG = (\Gamma, E)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{monotonicity assumption} iff:
%\[
%  \forall b \in \Gamma, \forall (A;B), (A';B') \in \GRNallres{b},
%  A \subset A' \wedge B' \subset B \Rightarrow K_{b,A,B} \leq_{[]} K_{b,A',B'}
%\]
$\forall b \in \Gamma$,
$\forall X^+ \subset \{ a \in \Gamma \mid \GRNedge{a}{+}{t}{b} \in E_+ \}$,
$\forall X^- \subset \{ a \in \Gamma \mid \GRNedge{a}{-}{t}{b} \in E_- \}$,
%A \subset A' \wedge B' \subset B \Rightarrow
$K_{b,\omega \cup X^-} \leq_{[]} K_{b,\omega \cup X^+}$
\end{property}

@@

\rewrite{This whole subsection may be completed, or turned into a new section}

\todo{Rewrite implementation with the new semantics}
